---
title: "cdc_findings"
output:
  pdf_document: default
  html_document: default
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Set Information:

This data set encompasses detailed statistics from the CDC's National Center for Health Statistics regarding live births in the United States, spanning from 2016 to 2022. It features a comprehensive range of demographic and health information derived from birth certificates. Key attributes include maternal age, pre-pregnancy BMI, birth weight, gestational age, county of residence, and various other health-related factors. This rich dataset offers insights into socio-economic, racial, and maternal and infant health aspects across different counties. The dataset, part of the National Vital Statistics System, is a valuable resource for in-depth analysis and research in public health and is accessible via the CDC WONDER Online Database.

## Objective: 
To understand the dataset's basic characteristics, identify patterns, outliers, anomalies, and underlying structures.

## Research Goal 
1.What are the key socio-economic and demographic predictors of maternal and infant health outcomes, as indicated by machine learning models?
2. Which machine learning algorithm provides the most accurate predictions of high-risk pregnancies based on socio-economic and racial factors?

## EDA - Exploratory Data Analysis
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(randomForest)
library(tidyverse)
library(tidyr)
library(caret)
library(corrplot)

# Read the data
data <- read.csv("cdc_county_data.csv")
head(data,5)


```

```{r}
# EDA: Descriptive Statistics
descriptive_stats <- summary(data)

# EDA: Distribution Analysis
pdf("distribution_analysis.pdf")
hist(data$Ave_Age_of_Mother, main="Distribution of Average Age of Mother", xlab="Average Age of Mother", col="blue", border="black")
hist(data$Ave_Pre_pregnancy_BMI, main="Distribution of Average Pre-pregnancy BMI", xlab="Average Pre-pregnancy BMI", col="green", border="black")
hist(data$Ave_Birth_Weight_gms, main="Distribution of Average Birth Weight (gms)", xlab="Average Birth Weight (gms)", col="red", border="black")
dev.off()

# EDA: Outlier Detection
pdf("boxplot_analysis.pdf")
boxplot(data$Ave_Age_of_Mother, main="Boxplot of Average Age of Mother", ylab="Average Age of Mother")
boxplot(data$Ave_Pre_pregnancy_BMI, main="Boxplot of Average Pre-pregnancy BMI", ylab="Average Pre-pregnancy BMI")
boxplot(data$Ave_Birth_Weight_gms, main="Boxplot of Average Birth Weight (gms)", ylab="Average Birth Weight (gms)")
dev.off()

# EDA: Initial Correlation Assessment
key_variables <- data[, c("Ave_Age_of_Mother", "Ave_Pre_pregnancy_BMI", "Ave_Birth_Weight_gms")]
correlation_matrix <- cor(key_variables, use="complete.obs")
corrplot(correlation_matrix, method="circle")

# Save the correlation plot
pdf("correlation_matrix.pdf")
corrplot(correlation_matrix, method="circle")
dev.off()

# Initial Visualization: Scatter Plot - Example
pdf("scatter_plot.pdf")
ggplot(data, aes(x=Ave_Age_of_Mother, y=Ave_Birth_Weight_gms)) +
  geom_point() +
  labs(title="Scatter Plot: Age of Mother vs Birth Weight", x="Average Age of Mother", y="Average Birth Weight (gms)")
dev.off()

```

### Descriptive Statistics
Ave_Age_of_Mother: Shows the average age of mothers in different counties and years.
Ave_Pre_pregnancy_BMI: Indicates the average Body Mass Index before pregnancy.
Ave_Birth_Weight_gms: Reflects the average birth weight of newborns in grams.

### Distribution Analysis
Maternal Age: The distribution of the average age of mothers seems roughly normal but may be slightly skewed towards younger ages.
Pre-pregnancy BMI: This variable also appears to follow a normal distribution, indicating a range of BMI values across the dataset.
Birth Weight: The birth weight distribution is roughly normal, suggesting a typical spread of newborn weights.

### Outlier Detection
Maternal Age: There are some outliers, indicating a few counties with unusually high or low average maternal ages.
Pre-pregnancy BMI: There are outliers present, suggesting some extremes in BMI values.
Birth Weight: Outliers are observed here as well, indicating some births with significantly high or low weights.


## Research Question 1: Key Socio-Economic and Demographic Predictors
We will use a machine learning model to identify the most significant predictors of maternal and infant health outcomes.
```{r}
# Install and load necessary libraries
if (!require("caret")) install.packages("caret")
library(caret)
if (!require("randomForest")) install.packages("randomForest")
library(randomForest)
if (!require("ROSE")) install.packages("ROSE")
library(ROSE)

# Selecting relevant predictors and the target variable
predictors <- data %>% select(Ave_Age_of_Mother, Ave_Pre_pregnancy_BMI, Ave_Birth_Weight_gms, Ave_Number_of_Prenatal_Wks, Maternal_Morbidity_YN)
target <- data$Births # Or any other relevant target variable

# Splitting data into training and test sets
set.seed(123)
trainIndex <- createDataPartition(target, p = .8, list = FALSE, times = 1)
trainData <- predictors[trainIndex, ]
testData <- predictors[-trainIndex, ]
trainTarget <- target[trainIndex]
testTarget <- target[-trainIndex]

# Training a Random Forest model
rf_model <- randomForest(trainData, as.factor(trainTarget), ntree = 100)
rf_importance <- importance(rf_model)

# Displaying feature importance
print(rf_importance)

```
Ave_Birth_Weight_gms (2429.82):
This variable, representing the average birth weight in grams, appears to be the most significant predictor in the model. A higher Mean Decrease Gini value indicates that this feature plays a crucial role in predicting the target variable.

Ave_Age_of_Mother (2290.50):
The average age of the mother is also a strong predictor, second only to birth weight. This suggests that maternal age significantly influences the outcome of interest.

Ave_Pre_pregnancy_BMI (2243.89):
The average pre-pregnancy BMI is another important predictor. Its influence is slightly less than that of maternal age but still substantial.

Ave_Number_of_Prenatal_Wks (2202.71):
The average number of prenatal weeks also shows considerable importance in the model. This reflects the impact of prenatal care duration on the target variable.

Maternal_Morbidity_YN (27.30):
This variable has a much lower importance score compared to others. It indicates whether maternal morbidity was noted, but it seems to have a lesser impact on the prediction of the target variable in the model.

note:
- The model suggests that birth weight, maternal age, pre-pregnancy BMI, and the number of prenatal weeks are critical factors.


## 2. ML Algorithm Accuracy

- using radial algorithm

```{r}
# Check data types before conversion
str(svm_trainData)

# Convert to numeric, handle non-numeric columns appropriately
svm_trainData <- data.frame(lapply(svm_trainData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))  # Convert factors or characters to numeric
  } else {
    as.numeric(x)  # Convert other types to numeric
  }
}))

# Ensure target variable is numeric
svm_trainTarget <- as.numeric(svm_trainTarget)

# Check again for NAs
sum(is.na(svm_trainData))
sum(is.na(svm_trainTarget))

# Verify the distribution of the target variable after sampling
table(svm_trainTarget)

# Retrain the SVM model
svm_model <- svm(x = svm_trainData, y = svm_trainTarget, kernel = "radial", cost = 0.5)

# Checkpoint 1: Model training completed
print("SVM model trained successfully.")

# Prepare test data for prediction
testData <- data.frame(lapply(testData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))
  } else {
    as.numeric(x)
  }
}))

# Checkpoint 2: Test data prepared for prediction
print("Test data prepared.")

# Make predictions
svm_prediction <- predict(svm_model, testData)

# Checkpoint 3: Prediction completed
print("Prediction completed.")

# Calculate RMSE
testTarget <- as.numeric(testTarget)  # Ensure testTarget is numeric
svm_rmse <- sqrt(mean((svm_prediction - testTarget)^2))

# Checkpoint 4: RMSE calculated
print("RMSE calculated.")

# Output RMSE
print(svm_rmse)

```

-using sigmoid

```{r}
# Check data types before conversion
str(svm_trainData)

# Convert to numeric, handle non-numeric columns appropriately
svm_trainData <- data.frame(lapply(svm_trainData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))  # Convert factors or characters to numeric
  } else {
    as.numeric(x)  # Convert other types to numeric
  }
}))

# Ensure target variable is numeric
svm_trainTarget <- as.numeric(svm_trainTarget)

# Check again for NAs
sum(is.na(svm_trainData))
sum(is.na(svm_trainTarget))

# Verify the distribution of the target variable after sampling
table(svm_trainTarget)

# Retrain the SVM model
svm_model <- svm(x = svm_trainData, y = svm_trainTarget, kernel = "sigmoid", cost = 0.5)

# Checkpoint 1: Model training completed
print("SVM model trained successfully.")

# Prepare test data for prediction
testData <- data.frame(lapply(testData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))
  } else {
    as.numeric(x)
  }
}))

# Checkpoint 2: Test data prepared for prediction
print("Test data prepared.")

# Make predictions
svm_prediction <- predict(svm_model, testData)

# Checkpoint 3: Prediction completed
print("Prediction completed.")

# Calculate RMSE
testTarget <- as.numeric(testTarget)  # Ensure testTarget is numeric
svm_rmse <- sqrt(mean((svm_prediction - testTarget)^2))

# Checkpoint 4: RMSE calculated
print("RMSE calculated.")

# Output RMSE
print(svm_rmse)

```

```{r}
# Check data types before conversion
str(svm_trainData)

# Convert to numeric, handle non-numeric columns appropriately
svm_trainData <- data.frame(lapply(svm_trainData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))  # Convert factors or characters to numeric
  } else {
    as.numeric(x)  # Convert other types to numeric
  }
}))

# Ensure target variable is numeric
svm_trainTarget <- as.numeric(svm_trainTarget)

# Check again for NAs
sum(is.na(svm_trainData))
sum(is.na(svm_trainTarget))

# Verify the distribution of the target variable after sampling
table(svm_trainTarget)

# Retrain the SVM model
svm_model <- svm(x = svm_trainData, y = svm_trainTarget, kernel = "polynomial", cost = 0.5)

# Checkpoint 1: Model training completed
print("SVM model trained successfully.")

# Prepare test data for prediction
testData <- data.frame(lapply(testData, function(x) {
  if (is.factor(x) || is.character(x)) {
    as.numeric(as.factor(x))
  } else {
    as.numeric(x)
  }
}))

# Checkpoint 2: Test data prepared for prediction
print("Test data prepared.")

# Make predictions
svm_prediction <- predict(svm_model, testData)

# Checkpoint 3: Prediction completed
print("Prediction completed.")

# Calculate RMSE
testTarget <- as.numeric(testTarget)  # Ensure testTarget is numeric
svm_rmse <- sqrt(mean((svm_prediction - testTarget)^2))

# Checkpoint 4: RMSE calculated
print("RMSE calculated.")

# Output RMSE
print(svm_rmse)

```

2. Which machine learning algorithm provides the most accurate predictions of high-risk pregnancies based on socio-economic and racial factors?
The radial algorithm did the best in rsme to train the model! This model assumption is worth looking into high pregnancies assumptions for a further deep dive.
